---
title: "Car Thefts"
---

Follow along with the demo to learn how to calculate standard deviation and how to create a histogram of the data using R and the Tidyverse. Get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```


### Task 2: Load the car thefts dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_jan24.csv


```{r}
car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_jan24.csv")
```

### Task 3: Calculate the standard deviation of the number of car thefts, along with the mean

```{r}
car_thefts |> summarize(mean = mean(count), sd = sd(count))
```


### Task 4: Create a histogram of the number of car thefts

```{r}
car_thefts |> 
  ggplot() + 
  geom_histogram(aes(x = count), binwidth = 2)
```




### Task 5: Add a line to the histogram that shows the mean of the number of car thefts, and two other lines to show one standard deviation in both directions


```{r}
car_thefts |> 
  ggplot() + 
  geom_histogram(aes(x = count), binwidth = 2) +
  geom_vline(aes(xintercept = mean(count)), color = "red", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(count) - sd(count)), color = "blue", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(count) + sd(count)), color = "blue", linetype = "dashed", size = 1)
```


### Task 6: Describe your results

Is the distribution normal or skewed in one direction? Write a sentence in which you describe the range of the data covered by one standard deviation from the mean on both sides - how much of the data falls in that range?


** The distribution is skewed further to the right, in this case, skewed to demonstrate that the majority of the dataset fell in between the mean and one standard deviation up. In simpler terms, it means that in our dataset, there were more dates in January of 2024 with between approximately 13 and 17 car thefts then there were dates with between approximately 8 and 12 car thefts. There were about 15 dates with between 13-17 car thefts, while there were about 9 dates with between 8-12 car thefts. **


When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
