---
title: "Home Values After Transit Line"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
options(scipen=999)
library(tidyverse)
```

## Measuring the Impact of Transit on Housing Prices

Local officials have claimed that a new light rail line has increased property values in a local neighborhood.

Before the transit line opened, homes in that neighborhood sold for an average of **$285,000**, with a standard deviation of **$32,000**.

Since the extension opened, your news team collected sales data from 14 recent home sales in the neighborhood:
$297,000, $305,000, $292,000, $318,000, $290,000, $311,000, $299,000, $325,000, $302,000, $288,000, $307,000, $315,000, $320,000, $294,000

As a journalist, you need to determine: **Is there statistical evidence that the transit line has actually increased home values? And if so, by how much?**

Follow along, running the code when directed. If you have questions, ask either in class or put them just below the section you have a question about.

### Task 1: Organize your data

```{r}
# Known information about home values before the transit line
prior_mean <- 285000  # average home value
prior_sd <- 32000     # standard deviation

# Home sale prices after the transit line opened (14 homes)
new_prices <- c(297000, 305000, 292000, 318000, 290000, 311000, 299000, 
                325000, 302000, 288000, 307000, 315000, 320000, 294000)

# Create a journalist-friendly dataset
home_data <- tibble(
  sale_id = 1:14,
  sale_price = new_prices
)

# View the data
home_data
```

### Task 2: Calculate key statistics (1 point)

Replace the REPLACE_ME values to calculate the mean, standard deviation and count of the new home prices.

```{r}
new_stats <- home_data |> 
  summarise(
    mean = mean(sale_price),
    sd = sd(sale_price),
    n = n()
  )

new_stats
```

**Answer this question:** How does the new average compare to the prior average of $285,000? How much higher is it in dollar terms? Does that seem like a big difference to you? What else would you want to know?

## The new average is $304,500 compared to the old average of $285,000, nearly a $20,000 increase. This doesn't seem like that big of a difference, but I believe the difference in the standard deviation is quite a big difference. The other things we should seek to know is size of the houses, how many incomes are paying for the houses, etc. ##

### Task 3: Visualize the distribution (1 point)

Let's make a histogram of the new home prices with a reference line showing the old average. This helps us see the spread of the data and where the prior average falls relative to the new sales.

```{r}
# Function to format currency values
dollar_format <- function(x) {
  paste0("$", format(x/1000, big.mark=","), "K")
}

ggplot(home_data, aes(x = sale_price)) +
  geom_histogram(binwidth = 5000, fill = "seagreen", color = "white", alpha = 0.8) +
  geom_vline(xintercept = prior_mean, color = "darkred", size = 1.2, linetype = "dashed") +
  geom_vline(xintercept = new_stats$mean, color = "darkblue", size = 1.2) +
  annotate("text", x = prior_mean - 1000, y = 3.5,
           label = "Previous Avg ($285K)", hjust = 1, fontface = "bold", color = "darkred") +
  annotate("text", x = new_stats$mean + 1000, y = 3,
           label = paste0("New Avg (", dollar_format(new_stats$mean), ")"), 
           hjust = 0, fontface = "bold", color = "darkblue") +
  scale_x_continuous(labels = dollar_format) +
  labs(
    title = "Home Sales After Transit Line Extension",
    subtitle = "Dashed line shows previous neighborhood average of $285,000",
    x = "Sale Price",
    y = "Number of Sales",
    caption = "Source: County Property Records"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10)
  )
```

**Answer this question:** Look at where the old average (red dashed line) falls relative to the new sales. How many of the 14 homes sold for more than the old average? What does the histogram tell you about the spread of the new prices?

## All of the 14 homes sold for more than the old average. The histogram tells us that the spread of the new prices is distributed somewhat evenly, but that it could very well be a significant difference. ##

### Task 4: Set up and perform a hypothesis test (2 points)

This is where we formally test the officials' claim. Fill in the REPLACE_ME values for the hypotheses and the t-test.

**Hypotheses:**

- **H₀ (Null):** The new light rail line is the reason why housing prices have increased.
- **H₁ (Alternative):** The new light rail line is not the reason why housing prices have increased. 

```{r}
# Set the significance level
alpha <- 285000

# Perform a one-sample t-test
# Since we want to know if home values INCREASED, we use alternative = "greater"
t_test_result <- t.test(
  home_data$sale_price,
  mu = 285000,
  alternative = "greater"
)

t_test_result
```

**Answer this question:** Look at the p-value. Is it less than our alpha of 0.05? What does this mean for our null hypothesis — can we reject it? What can we now say about the officials' claim?

## Yes, the p-value is less than our alpha of 0.05. This means that we cannot reject the null hypothesis and that we can't say that the officials' claim that housing prices have increased due to the light rail line is due to random chance. ##

### Task 5: Interpret the results with a confidence interval (2 points)

Remember confidence intervals from last week? The t-test gives us one too. Let's pull together the key statistics for your story.

```{r}
# Extract the p-value
p_value <- t_test_result$p.value

# Calculate the 95% confidence interval (two-sided for the full range)
ci <- t.test(home_data$sale_price)$conf.int

# Key statistics for your story
story_stats <- tibble(
  `Previous average` = dollar_format(prior_mean),
  `New average` = dollar_format(mean(new_prices)),
  `Increase` = dollar_format(mean(new_prices) - prior_mean),
  `Percent change` = paste0(round(((mean(new_prices) - prior_mean) / prior_mean) * 100, 1), "%"),
  `p-value` = p_value,
  `95% CI lower` = dollar_format(ci[1]),
  `95% CI upper` = dollar_format(ci[2]),
  `Significant?` = if_else(p_value < alpha, "Yes", "No")
)

story_stats
```

**Answer this question:** The 95% confidence interval tells us the range where the true average home price likely falls. Does that range include the old average of $285,000? What does that tell you?

## The range of the 95% confidence interval does not include the old average, meaning that the change is not only not random, but also significant. ## 

### Task 6: Calculate the effect size (3 points)

We've established that the increase is statistically significant — it's real, not due to chance. But how big is it, really? We'll calculate Cohen's d to find out. The formula is:

**d = (new mean - prior mean) / prior standard deviation**

Replace the REPLACE_ME values.

```{r}
effect_size <- (mean(new_prices) - prior_mean) / prior_sd

effect_size_label <- case_when(
  abs(effect_size) < 0.2 ~ "Negligible",
  abs(effect_size) < 0.5 ~ "Small",
  abs(effect_size) < 0.8 ~ "Medium",
  TRUE ~ "Large"
)

tibble(
  `Cohen's d` = round(effect_size, 3),
  `Interpretation` = effect_size_label,
  `Dollar increase` = dollar_format(mean(new_prices) - prior_mean),
  `Percent increase` = paste0(round(((mean(new_prices) - prior_mean) / prior_mean) * 100, 1), "%")
)
```

**Answer this question:** The effect size is "medium." In dollar terms, the average increase is about $19,500. But remember that the standard deviation of home prices in this neighborhood was $32,000 *before* the transit line — meaning prices already varied by tens of thousands of dollars from sale to sale. Given that context, does a $19,500 increase seem large or modest? How does the effect size help you think about this differently than just looking at the dollar amount?

## The increase of $19,500 is modest, given our previous mean. Once again, I believe that the discrepency between the old and new standard deviations is more significant. The effect size of the mean helps paint a more clear picture than just looking at the dollar amounts would: $19,500 is a decently big number, but not necessarily in relation to the costs of houses. ##

### Task 7: What's the story? (6 points)

You now have all the statistical evidence. But writing the story requires more than numbers — it requires judgment.

**Part A:** The local officials are going to claim the transit line increased home values. Based on your analysis, is that claim supported? Write 2-3 sentences for the top of your story that accurately reflect both the statistical significance AND the effect size. Don't just echo the officials' talking points — interrogate them.

## Based on the data analysis, the officials' claim that the transit line increased housing prices is supported - it wasn't random. However, the analysis also suggests that the improvement, while statistically significant, was a modest increase rather than a dramatic one. The new transit line hasn't made a large difference in the prices that houses have sold for. ##

**Part B:** A $19,500 increase is real and statistically significant, but what *else* could explain rising home prices in this neighborhood? List at least two alternative explanations that a good reporter would want to investigate before attributing the increase entirely to the transit line.

## Alternative explanations for why home prices have risen include: perhaps the housing market has improved or become more stable. Perhaps there was another thing built in the community that made it more desirable: a school, a new business headquarters, etc. These are things that a good reporter would want to investigate first before entirely attributing the rise to the transit line. ##


When you are finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.