---
title: "JOUR405: Statistics for Journalists"
subtitle: "Midterm Exam"
name: "Andrew Breza"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# JOUR405: Statistics for Journalists
## Midterm Exam

For this exam, you'll analyze several datasets using R and the statistical concepts we've covered in class. Load the tidyverse before beginning, then complete each task. Write your code in the provided blocks and answer the questions in complete sentences.

## Load the tidyverse

```{r}
library(tidyverse)
```

---

## Part 1: Restaurant Health Inspections (25 points)

The first dataset contains restaurant health inspection scores from Montgomery County. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv`

The dataset includes the name of the establishment, the number of points for critical and non-critical areas, the total points, maximum points possible, the compliance score, and the grade.

### Task 1: Read the data (2 points)

```{r}
moco_inspections <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv")
```

### Task 2: Calculate the mean, standard deviation, minimum, and maximum of compliance scores. Save the results in a new dataframe called `compliance_stats`. (2 points for each)

```{r}
compliance_stats <- moco_inspections |>
mutate(
  mean_score = mean(compliance_score),
  sd_score = sd(compliance_score),
  min_score = min(compliance_score),
  max_score = max(compliance_score)
)
```


**Mean: 96.30665**
**Standard Deviation: 5.826279**
**Minimum: 42**
**Maximum: 100**

### Task 3: Create a histogram of the compliance scores with a vertical line showing the mean. Include a descriptive title and label the x and y axes using `labs()`. (10 points)

```{r}
ggplot() +
  geom_histogram(data = compliance_stats, aes(x = compliance_score)
  geom_vline(xintercept = 96.30665),
  labs(x = Number of places, y = Overall compliance score))
```

### Task 4: Write 2–3 sentences interpreting what the standard deviation and histogram tell us about the distribution of compliance scores. What is newsworthy, and what would your next reporting step be? (5 points)

**Answer:**

The most newsworthy thing is that this data set is very far from being normally distributed. It has extreme outliers that greatly skew any sort of perception and subsequent story.

---

## Part 2: Daily Transit Ridership — Z-Scores and Sampling (30 points)

Your editor wants you to identify unusual days in WMATA's daily ridership data that might be worth investigating as story leads. The dataset contains daily bus and rail totals. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv`

### Task 5: Load the WMATA ridership data. (2 points)

```{r}
wmata_daily <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv")
```

### Task 6: Using `summarize()`, calculate the mean and standard deviation for both bus and rail ridership. Save the results in a dataframe called `wmata_stats`. (5 points)

```{r}
wmata_stats <- wmata_daily |>
summarise(
  mean_bus = mean(bus),
  sd_bus = sd(bus),
  mean_rail = mean(rail),
  sd_rail = sd(rail)
)
```


**Mean_Bus: 331293.8**
**SD_Bus: 88486.88**
**Mean_Rail: 347987.8**
**SD_Rail: 100164.3**


### Task 7: Using `mutate()`, add two new columns to `wmata_daily`: `rail_zscore` (the z-score for rail ridership) and `bus_zscore` (the z-score for bus ridership). Save the result as `wmata_zscores`. (8 points)

```{r}
wmata_zscores <- wmata_daily |>
mutate(
  bus_zscore = (total - 331293.8)/88486.88 * 100,
  rail_zscore = (total - 347987.8)/100164.3 * 100
)
```

### Task 8: Filter `wmata_zscores` to show only days where the absolute value of `rail_zscore` is greater than 2 (unusually high or low rail ridership). Arrange by the absolute value of the z-score, with the most extreme first. How many unusual days are there? (8 points)

```{r}
wmata_zscores |>
filter(rail_zscore case_when(abs(>2)))
```

### Task 9: Using `set.seed()`, take a random sample of 91 days from `wmata_daily` (roughly 25% of the data). Calculate the mean and standard deviation for bus and rail ridership in the sample. How do the sample statistics compare to the population statistics you calculated in Task 6? Write 1–2 sentences comparing the values and explaining what this suggests about random sampling. (4 points)

```{r}
set.seed(91)
```

### Task 10: Write 2–3 sentences: What do the unusual days in your z-score analysis suggest as story leads? (3 points)


---

## Part 3: Maryland Car Theft Rates (25 points)

Your editor has assigned you a story about car thefts in Maryland and wants you to find out which counties have the highest rates. The dataset contains car theft counts for Maryland counties in 2023 along with population figures. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv`

### Task 11: Load the car thefts data. (2 points)

```{r}
md_car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv")
```

### Task 12: Using `mutate()`, add a column called `theft_rate` that calculates the car theft rate per 10,000 residents for each county. Save the result back to `md_car_thefts`. In 1–2 sentences, explain why a per-capita rate is more useful than raw counts for this comparison. (8 points)

```{r}
md_car_thefts |>
mutate(
  theft_rate = population/10000
)
```


**A per capita rate would be better here because Maryland counties don't have even populations, so we can expect that the counties with higher populations will have more car thefts. A per capita rate would show us whicbh countries truly do haved a car theft issue.**

### Task 13: Calculate the median car theft rate and the total number of car thefts statewide. Show the counties that have rates above the median. (7 points)

```{r}
```

### Task 14: Write 2–3 sentences describing what these calculations reveal about the distribution of car thefts in Maryland. What is the lede of a story based on your findings? (8 points)

---

## Part 4: Data Analysis Scenario (20 points)

A city parks department releases data showing that its 12 community swimming pools average 187 visitors per day during the summer season, with a standard deviation of 42 visitors. Your editor suspects that a handful of pools are dramatically underperforming and wants a story. You also learn that the department's figures are based on a random sample of 60 days per pool rather than a complete count.

Write 3–4 sentences for each of the following questions:

### Question A: Given that the mean is 187 and standard deviation is 42, which statistical measures would you prioritize using and why? (7 points)


**The statistical measures that I would prioritize using would be the Central Limit Theorem and also considering creating a poll to gain more accurate feedback from the community members attending the pool. These methods would be useful in making a potential story more accurate numerically and logically since we could prove numbers with more certainty and have actual testimony from the community involved.**


### Question B: How would you use z-scores to identify which pools are genuinely underperforming, and why is that more useful than simply ranking pools by raw attendance? (7 points)


**We would use Z-scores to identify which pools are underperforming by figuring out each pool's specific mean and standard deviation and compare it to those of the 12 pools as a whole, what we were given. By doing this, we can see each pool's Z-scores on average, and if we wanted to, could do a day-by-day comparison. This is a better tool than using raw attendance because raw attendance is more easily swayed by outliers. Perhaps one pool had a massive party one day but was relatively empty most other days. That party would heavily sway their raw attendance.**


### Question C: The department collected its attendance figures using a random sample of 60 days per pool rather than tracking every day. What questions would you ask about the sample before relying on these figures for a story? What concerns would you raise in your story? Use statistical terminology where appropriate. (6 points)


**Some questions I would ask about the sample would include:
What was the highest and lowest recorded temperature on those 60 days?
Were there any major holidays within those 60 days?
Do any of these pools require payment or passes to enter?**

**The main concern that I would raise in my story would be that there are many different factors that go into attendance at a pool, so these numbers may tell part of a story, but not necessarily the whole story.**


---

### Submission Instructions
- Save your work frequently
- Make sure all code blocks run without errors
- Provide clear explanations for your choices

Good luck!
