---
title: "Identifying Schools with Biggest Differences in Math Test Scores, Part 2"
author: "Andrew Breza"
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

### Introduction

In the class activity, you tested a school board member's claim using data from Montgomery County, Pa. — and found a model that didn't hold up. Now you'll run the same analysis using data from Montgomery County, Md. Your job is to evaluate whether the Maryland data tells a different story, and what that means for how you'd report on the claim.

### Loading the Data

```{r}
schools <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/montgomery_md.csv")
```

### Task 1: Examine the Data First (2 points)

Before you plot anything, look at the data — you know why by now.

```{r}
summary(schools$`FARMS Pct`)
schools |> arrange(desc(`FARMS Pct`)) |> head(10)
```

Are there any data quality issues like the ones you found in the Pennsylvania data? How does the distribution of FARMS percentages compare between the two counties? Note anything that concerns you or gives you confidence in the data.

## The data in this Maryland example seems a lot more quality than the data we looked at in Pennsylvania - there are no schools in this example with a 100% FARMS percentage, while there were eight in Pennsylvania. In the latter, there were also none between 100% and 54%, while we have lots of schools between those two figures in Maryland. Right off the bat, this gives me more confidence in our ability to prove the null hypothesis. ##


### Task 2: Create a Scatterplot and Compare (5 points)

Replace the REPLACE_ME value with the correct variable name to create a scatterplot of FARMS percentage vs. math scores for Montgomery County, Md.

```{r}
schools |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations",
       title = "Math Scores vs. FARMS Percentage, Montgomery County MD")
```

Describe what you see. Then compare this scatterplot specifically to the Pennsylvania one: what differences do you notice in the direction, strength, and spread of the relationship? The two counties share a name and are both suburban — what might explain why they look so different?

## Based on looks, it appears that there is a decent correlation between lower test scores and higher FARMS percentage. At the very least, it's far more clear than the scatterplot for the Pennsylvania schools in direction, strength, and spread of the data. Factors that might explain the differences between the two counties include a less diverse wealth disparity, more resources for schools in the county, etc. ##


### Task 3: Linear Regression Model (5 points)

Replace the REPLACE_ME value with the correct variable name to run a linear regression model predicting math scores from FARMS percentage.

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)
summary(model)
```

Report the three key statistics: the p-value, the Adjusted R-squared, and the coefficient for FARMS Pct. Then interpret them together — not separately. What does the combination of these three numbers tell you about the relationship between poverty and math scores in Montgomery County, Md.? Is this model worth using? How does it compare to the Pennsylvania model?

## All three values - the p-value, Adjusted R-squared, and the coefficient - suggest that there is a strong relationship between poverty and math scores in Montgomery County, Md., and that this model is worth using. Our p-value is statistically significant, the Adjusted R-squared value could be better but explains a good amount of the variance, and the coefficient suggests that for every increase of 1 in FARMS Pct, math scores drop by .59. All of these numbers exceed the data in Montgomery County, Pa.: the p-value is not statistically significant, the Adjusted R-squared value only accounts for about 2% of the variance, and for each increase of 1 in FARMS Pct, test scores rise by .12. ##


### Task 4: Identifying Schools with the Biggest Residuals (6 points)

Replace the REPLACE_ME values with the correct variable names to calculate predicted values, residuals, and absolute residuals for each school. Then arrange the schools in descending order of absolute residuals.

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)
summary(model)
```


```{r}
schools <- schools |>
  mutate(
    predicted = predict(model),
    residual = `% Met or Exceeded Math Expectations` - predicted,
    abs_residual = abs(residual)
  ) |>
  arrange(desc(abs_residual))

head(schools |> select(1, `FARMS Pct`, `% Met or Exceeded Math Expectations`, predicted, residual), 10)
```

Identify the school with the largest positive residual (doing much better than the model predicts) and the school with the largest negative residual (doing much worse). For each school, find its zip code and look it up on [CensusReporter.org](https://censusreporter.org).

## The school with the largest positive residual is Fox Chapel Elementary, while the school with the largest negative residual is Germantown Elementary. They're both located in Germantown, but with slightly different zip codes (20876 vs 20874, respectively). ##

For each of the two schools, answer:
- What does the demographic and economic profile of the zip code look like?
- Does anything in that profile help explain why the school is over- or under-performing the model?
- What would your first reporting question be to the principal of that school?

## Comparing 20874 and 20876, the former has a much higher population, a lower median household income, and a slightly lower percentage of people below the poverty line. These figures tell quite a bit in terms of difference in math scores, as Germantown High School is the one in 20874. However, it doesn't support the FARMS Pct angle, as Fox Chapel has a decently higher percentage of students on FARMS (56.7 versus 35.9). The first question to the principal might be centered around what sort of resources the county versus city is providing them. ##


### Task 5: Evaluate the Model and Write the Pitch (7 points)

You've now run this analysis on two counties. Return to the school board member's claim: that elementary schools with fewer students on free and reduced-price meals do better on standardized math tests.

Write a short story pitch of 4-6 sentences that addresses all of the following: Does the Maryland data support or contradict the claim? How does it compare to what you found in Pennsylvania? What would you ask the school board member if you called them for comment? And what's the most interesting specific finding from your residuals analysis: which school would you want to visit first, and why?

## The Maryland data supports the initial claim that schools with fewer students on FARMS tend to perform better on standardized math tests, but the Pennsylvania data does not. An analysis of the Maryland data showed a much higher correlation and more concrete evidence to back the claim up. If I were to call the board member, especially as somebody far more familiar with Maryland's Montgomery County, I'd like to know what else they might attribute to lower standardized math test scores. In terms of the school I'd like to visit, it would be Germantown Elementary School, since they are not anywhere near the bottom or the top of FARMS percentages yet are underperforming at the highest rate. What other causes might there be? ## 