---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2025.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

**It's very strange to me that a number of employees with titles including "senior" or "assistant/advisor to the President" have no reported salary. That's 100% something that should be investigated further, especially since the highest paid employee, sitting at $225,700, is labled as just "Advisor."**

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries |>
summarize(
  avg_salaries = mean(SALARY),
  median_salaries = median(SALARY)
)
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries |>
summarize(
  mean_salary = mean(SALARY),
  median_salary = median(SALARY),
  min_salary = min(SALARY),
  max_salary = max(SALARY)
  )
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

**The mean and median salaries show us that the vast majority of white house employees are incredibly well-paid. The minimum and maximum value shows that there are certain positions that are much more highly regarded through their level of income. Both the minimum and maximum salaries are extreme outliers - the next lowest to zero is $59,070 and the next highest to $225,700 is $203,645.**

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

**My story idea would be an investigative piece into the specific tasks that some of these employees undertake, with a focus on the outliers. The next step to take in that would be to pitch it to an editor and then send emails to the respective employees (or perhaps the white house's pr department) to try to set up an interview.**