---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library(tidyverse)
library(janitor)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv")
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities_rates <- md_cities |>
  mutate(
    violent_rate_2020 = violent2020/pop2020 * 1000,
    property_rate_2020 = property2020/pop2020 * 1000,
    violent_rate_2019 = violent2019/pop2019 * 1000,
    property_rate_2019 = property2019/pop2019 * 1000
  )
```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?


** To begin, I decided to make the rate 1000 since there were only 39 of the 137 cities with a population of less than 1000. Trying to do 100 just to include them doesn't yield any good findings. **

** One interesting finding was that of the six cities with a double digit violent crime rate in both years, four of them grew in that measure in 2020 compared to 2019. Baltimore and Williamsport are the two that shrunk, which could potentially be explained by the pandemic, but if they shrunk, why did the others rise? Especially Ocean City, which rose from 13 to 23. Ocean City was also the only city with a triple digit rate in property crime rate, which also rose significantly from 2019 to 2020. **

### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change <- md_cities_rates |> 
  mutate(
    violent_pct_change = (violent_rate_2020 - violent_rate_2019)/violent_rate_2019 * 100,
    property_pct_change = (property_rate_2020 - property_rate_2019/property_rate_2019 * 100)
  )
```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

** The city of Edmonston immediately stands out. What caused an 895% increase in the violent crime rate? And, once again, Ocean City stands out as it is the only city to have an increase in property crime rate year over year. I think these examples show that the outliers are far more interesting though. **

### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

** Nine cities in Maryland saw over a 100% increase in violent crime rate from 2019 to 2020.**

** While four of the cities are located in the two most populous counties in Maryland, Montgomery and Prince George's, four cities are located amongst Worcester and Caroline Counties, both in the bottom eight in population amongst Maryland's 24 counties. **

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
