---
title: "HW7b - Probability and the Normal Distribution"
---

In this assignment, you'll build on your z-score skills to answer probability questions using the normal distribution. You'll return to the White House salaries data and use it to answer the kinds of "what are the chances?" questions that come up in journalism all the time. This sets the stage for sampling, which we'll cover next.

### Task 1: Load the tidyverse and the data (1 point)

Load the tidyverse and the White House salaries data from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv into a dataframe called `wh_salaries`

```{r}
library(tidyverse)
```

```{r}
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv")
```


### Task 2: Filter and summarize (2 points)

You may remember from HW3 that some White House staffers have a salary of $0. Let's filter those out and focus on paid employees, saving the result to a new dataframe called `wh_paid`. You need to calculate the mean, median and standard deviation of the salary column using `wh_paid`:

```{r}
wh_paid <- wh_salaries |>
  filter(SALARY > 0)

wh_paid |>
  summarise(
    mean_salary = mean(SALARY),
    median_salary = median(SALARY),
    sd_salary = sd(SALARY)
  )

```

Now create a histogram of the paid salaries with a vertical line for the mean and two more for ±1 standard deviation, similar to what you did in HW6 with Spotify data.

```{r}
wh_paid |>
  ggplot() +
  geom_histogram(aes(x=SALARY))
```


```{r}
wh_paid |> 
  ggplot() + 
  geom_histogram(aes(x = SALARY)) +
  geom_vline(aes(xintercept = mean(SALARY)), color = "red", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(SALARY) - sd(SALARY)), color = "blue", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(SALARY) + sd(SALARY)), color = "blue", linetype = "dashed", size = 1)
```

**Describe the shape of the distribution. Is it roughly normal? Skewed? Where do most salaries cluster?**

** Our distribution is hardly normal. It's very erratic. Most salaries cluster around the first standard deviation lower than the mean, but there are also a decent chunk slightly higher than the mean, while another decent chunk is around $200,000. **

### Task 3: The Empirical Rule (3 points)

The empirical rule (also called the 68-95-99.7 rule) says that for a normal distribution:
- About 68% of data falls within 1 standard deviation of the mean
- About 95% falls within 2 standard deviations
- About 99.7% falls within 3 standard deviations

Let's test this. We'll calculate the salary ranges for 1, 2, and 3 standard deviations from the mean, then count how many employees actually fall in each range.

```{r}
mean_sal <- mean(wh_paid$SALARY)
sd_sal <- sd(wh_paid$SALARY)
total_n <- nrow(wh_paid)

empirical <- tibble(
  range = c("Within 1 SD", "Within 2 SD", "Within 3 SD"),
  lower = c(mean_sal - sd_sal, mean_sal - 2 * sd_sal, mean_sal - 3 * sd_sal),
  upper = c(mean_sal + sd_sal, mean_sal + 2 * sd_sal, mean_sal + 3 * sd_sal),
  expected_pct = c(68, 95, 99.7),
  actual_count = c(
    sum(wh_paid$SALARY >= (mean_sal - sd_sal) & wh_paid$SALARY <= (mean_sal + sd_sal)),
    sum(wh_paid$SALARY >= (mean_sal - 2 * sd_sal) & wh_paid$SALARY <= (mean_sal + 2 * sd_sal)),
    sum(wh_paid$SALARY >= (mean_sal - 3 * sd_sal) & wh_paid$SALARY <= (mean_sal + 3 * sd_sal))
  ),
  actual_pct = round((actual_count / total_n) * 100, 1)
)

empirical
```

**How closely does the data follow the empirical rule? What does any deviation tell you about the distribution?**

** The data is decently far away from following the empirical rule. None of the actual percentages are particularly close to the expected percentages, outside of the +/- 3 standard deviations. This shows us that our data is not normally distributed. **

### Task 4: From z-scores to probability with pnorm() (4 points)

Remember the car thefts z-score activity from class? Well, I want to return to that briefly, introduce a new tool called pnorm() and then apply it here. So open that CarTheftZScores notebook and, at the bottom, copy and paste in the contents here: https://gist.githubusercontent.com/dwillis/25102889f576b1707576b85e0be0c353/raw/1ca2be9636497cb5293ff7b1f4e6ed8e27a9e049/car_theft_pnorm.rmd

Re-run all of the blocks in CarTheftZScores notebook, including the code you just pasted in. What's happening there is you are using a function called `pnorm()` to find the probability of a value being less than or equal to a certain z-score. The basic formula to find that probability is:

```
1 - pnorm(z_score)
```

Where `z_score` is one row's z-score. Now apply the same approach to salaries.

Your editor asks: **"What's the probability that a randomly selected White House employee makes more than $150,000?"**

To answer this:
1. Calculate the z-score for $150,000
2. Use `pnorm()` to find the probability. `pnorm()` gives you the area under the normal curve to the LEFT of a z-score — that is, the probability of being BELOW that value.
3. Subtract from 1 to get the probability of being ABOVE $150,000
4. Check your answer against the actual data

```{r}

salary <- 150000
z_score <- (salary - mean_sal) / sd_sal

prob_less <- pnorm(z_score)
prob_more <- 1 - prob_less
prob_more

# Check against actual data
actual_prob <- sum(wh_paid$SALARY > 150000) / total_n
actual_prob
```

**Write a sentence describing this probability in plain English. How does the normal approximation (prob_less) compare to the actual percentage?**

** The approximation of selecting a random White House employee who makes more than $150,000 is around 16.88%, while the actual percentage of White House employees who make more than that figure is 18.18%. This tells us that there's about an 83% chance that the random employee chosen will not be one who makes more than $150,000 **

### Task 5: Probability between two values (3 points)

Now your editor asks: **"What percentage of White House employees make between $80,000 and $120,000?"**

You'll need to calculate z-scores for those two figures, use `pnorm()` for each, and subtract to find the probability between them. Update the REPLACE_ME figures with the appropriate numbers.

```{r}

z_lower <- (80000 - mean_sal) / sd_sal
z_upper <- (120000 - mean_sal) / sd_sal

prob_between <- pnorm(z_upper) - pnorm(z_lower)
prob_between

# Verify with actual data
actual_between <- sum(wh_paid$SALARY >= 80000 & wh_paid$SALARY <= 120000) / total_n
actual_between

```

**Interpret this result. What does this "middle range" tell us about the White House salary structure?**

** Based on the distribution of the data, we would expect 36.64% of White House employees to make between 80,000 and 120,000 dollars, when the actual number of employees in that range is 32.58%. **

### Task 6: Expected vs. observed (3 points)

Let's compare what the normal distribution *predicts* to what we actually *see*. This idea — comparing expected to observed — is one you'll use again later in the course.

Create these salary bins and calculate the expected count (from the normal distribution) and the observed count (from the actual data) for each:

```{r}
bins <- tibble(
  category = c("Below $60k", "$60k-$90k", "$90k-$120k", "$120k-$150k", "Above $150k"),
  lower = c(-Inf, 60000, 90000, 120000, 150000),
  upper = c(60000, 90000, 120000, 150000, Inf)
) |>
  mutate(
    expected_pct = pnorm(upper, mean = mean_sal, sd = sd_sal) - pnorm(lower, mean = mean_sal, sd = sd_sal),
    expected_count = round(expected_pct * total_n, 1),
    observed_count = c(
      sum(wh_paid$SALARY < 60000),
      sum(wh_paid$SALARY >= 60000 & wh_paid$SALARY < 90000),
      sum(wh_paid$SALARY >= 90000 & wh_paid$SALARY < 120000),
      sum(wh_paid$SALARY >= 120000 & wh_paid$SALARY < 150000),
      sum(wh_paid$SALARY >= 150000)
    ),
    difference = observed_count - expected_count
  )

bins |> select(category, expected_count, observed_count, difference)
```

**Where does reality differ most from what the normal distribution predicts? What might explain those differences?**

** Reality differs most for the employees making less than 60K and those between 60-90K. A normal distribution predicts around 40 employees to make less than 60K, while only two actually do, a difference of -38. For those between 60-90K, the prediction is 78 employees, while there are actually 151, a difference of 73. The expected answer for why there are such vast disparities is job title, but the titles we're given are all relatively similar. We'd have to do more research into what specific functions are executed by the employees in each bin. **

### Task 7: Story pitch (4 points)

Based on your probability analysis, write a headline and lede for a story about White House salaries. Your lede should incorporate at least one probability or percentage you calculated.

Then write 2-3 sentences describing how you would use probability thinking to advance this story — what could you compare these probabilities to, and what additional analysis would help?

** Headline: Salaries for White House employees are all over the place - but why? **

** Lede: Statistical analysis shows that only about 32% of White House employees make between 80000 and 120000 dollars each year. Where do the rest fall, and why are the salaries so far apart? **

** Probability thinking advances this story because it shines a light on something that's quite important in the United States. The ethos of journalism is grounded in accountability, and through this research, something seems very off about why these salaries are all over the place. You could compare these salaries to other important government salaries such as Congress, the Pentagon, etc. Additional analysis would include exploring specific job functions and how the numbers within different government agenices compare. **

### Task 8: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.